---
title: "California Housing Price Prediction: A case study"
subtitle: "Moving beyond linearity"
author: "Group 1"
date: "2026-05-18"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true
    theme: flatly
    highlight: tango
    number_sections: true
    code_folding: show
    df_print: paged
    fig_width: 10
    fig_height: 6
---

**Team Members:**
Collins Oduor - 219858 | Michelle Kalevela - 222647 | Diana Otieno - 222658 |
Methuselah Amtany - 223930 | Wesley Gathua - 223347 | Dennis Koome - 193582 |
Wendy Oira - 221007 | Nehema Mburuki - 190546

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 10,
  fig.height = 6,
  comment = "#>"
)
```

# Introduction & Objectives

This project applies the nonparametric and semi-parametric regression techniques to the California Housing dataset. The central thesis is that the relationship between housing price and its predictors is **inherently nonlinear and spatially interactive**, a claim we test rigorously by comparing a full ladder of model families.

**Research Questions**

1. Can ensemble nonlinear models (Random Forest, XGBoost, Neural Network) predict house prices more accurately than classical approaches? What are the most important nonlinear interactions between location and income?

2. Which model family — **GNPR** (GAM, Kernel/Local Regression, Splines, Step Functions, Polynomial) vs. **Parametric** (Linear, Ridge) vs. **Machine Learning** (RF, XGBoost, NN) — provides the best predictive performance?

# Package Installation & Loading

```{r packages}
required_pkgs <- c(
  "tidyverse", "caret", "randomForest", "xgboost",
  "mgcv", "splines", "np", "corrplot", "gridExtra",
  "Metrics", "neuralnet", "ggplot2", "viridis", "scales",
  "pdp", "vip", "spdep", "sf", "kableExtra", "GGally",
  "patchwork", "ggrepel", "finalfit", "spatialreg", "spdep",
  "sf", "spam", "ranger", "GWmodel", "tmap"
)

for (pkg in required_pkgs) {
  if(!(pkg %in% installed.packages())) {
    install.packages(pkg)
  }
  
  library(pkg, character.only = T)
}

# Global ggplot theme
theme_set(theme_minimal(base_size = 12) +
            theme(plot.title = element_text(face = "bold", size = 13),
                  plot.subtitle = element_text(color = "grey40")))
```

---

# Constants and Utility functions

```{r}
columns_all <- c("longitude", "latitude", "housing_median_age",
                 "total_rooms", "total_bedrooms", "population",
                 "households", "median_income", "median_house_value")
columns_non_spatial <- c("housing_median_age","total_rooms", 
                         "total_bedrooms", "population",
                         "households", "median_income", 
                         "median_house_value")
independent_columns_all <- c("longitude", "latitude", 
                             "housing_median_age","total_rooms", 
                             "total_bedrooms", "population",
                             "households", "median_income")
independent_columns_non_spatial <- c("housing_median_age",
                                     "total_rooms", "total_bedrooms", 
                                     "population", "households", 
                                     "median_income")

set.seed(42)
```

```{r helper-functions}
calc_metrics <- function(actual, predicted, model_name) {
  rmse_val <- rmse(actual, predicted)
  mae_val  <- mae(actual, predicted)
  r2_val   <- cor(actual, predicted)^2
  tibble(
    Model = model_name,
    RMSE  = round(rmse_val, 2),
    MAE   = round(mae_val,  2),
    R2    = round(r2_val,   4)
  )
}

# Storage for predictions and metrics
all_metrics <- list()
all_preds   <- list()
```

---

# Data Loading & Initial Inspection

```{r load-data}
data <- read.csv("~/Downloads/cal_housing.data", header = FALSE)

colnames(data) <- columns_all

cat("Dimensions: ", dim(data), "\n")
cat("Missing values: ", sum(is.na(data)), "\n")

head(data, 6)
```

```{r data-summary}
cat("Summary of data \n")
t(summary(data))

cat("\nUnique values per column \n")
t(sapply(data, function(x) length(unique(x))))
```

---

# Exploratory Data Analysis

## Distributions of All Variables

```{r eda-distributions, fig.height=8}
# 1. Calculating the medians for each variable first
medians_data <- data %>% 
  select(-c(longitude, latitude)) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
  group_by(variable) %>%
  summarize(median_val = median(value, na.rm = TRUE))

# 2. Pivoting the main data
data_long <- data %>% 
  select(-c(longitude, latitude)) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value")

# 3. Plotting then adding the medians_data to the geom_vline
ggplot(data_long, aes(x = value)) +
  geom_histogram(bins = 40, fill = "steelblue", color = "white", alpha = 0.8) +
  geom_vline(data = medians_data, aes(xintercept = median_val), color = "red", linetype = "dashed", size = 1) +
  facet_wrap(~variable, scales = "free") +
  theme_minimal() +
  labs(title = "Distribution of Variables", x = "Value", y = "Frequency")
```

- The numeric features are skewed to the right apart from housing_median_age which is bi-modal.


```{r checking the max values of the capped variables}
cat("The highest value for housing median age is:", max(data$housing_median_age), "\n")
cat("The highest median house value is:", max(data$median_house_value), "\n")
cat("The highest median income is:", max(data$median_income), "\n")
```
- Housing_median_age is capped at 52 years, median_house_value at approximately \$500,000 and median_income at \$15,000. Since the values at the cap represent elements whose true amounts are higher, they were filtered out to avoid corrupting the model using unreal data.

```{r, filtering capped values}
# Creating a clean, uncapped dataset
clean_data <- data %>%
  filter(median_house_value < 500000) %>%
  filter(housing_median_age < 52) %>%
  filter(median_income < 15)

# Checking # of lost rows
print(paste("Original rows:", nrow(data)))
print(paste("Cleaned rows:", nrow(clean_data)))
```
- After filtering, 2087 rows of data were lost. A sacrifice for statistical accuracy.

## Correlation Structure

```{r eda-correlation, fig.height=7}
cor_matrix <- cor(clean_data, use = "complete.obs")
corrplot(
  cor_matrix,
  method      = "color",
  type        = "upper",
  tl.cex      = 0.85,
  addCoef.col = "black",
  number.cex  = 0.60,
  col         = colorRampPalette(c("#d73027","white","#1a6aaa"))(200),
  title       = "Pearson Correlation Matrix",
  mar         = c(0, 0, 2, 0)
)
```

- There is a very high correlation between total bedrooms, population, households and total rooms. Median income has a high correlation with median house value.

```{r finalfit-summary-table}
# 1. Creating categorical dependent variable using existing 'clean_data'
summary_data <- clean_data %>%
  mutate(
    house_value_group = ntile(median_house_value, 4) %>%
      factor(labels = c("Low", "Medium", "High", "Very High"))
  )

# 2. Defining dependent and explanatory variables
dependent <- "house_value_group"

# Selecting all columns except targets and geographical coordinates
explanatory <- summary_data %>%
  select(-median_house_value, -house_value_group, -longitude, -latitude) %>%
  names()

# 3. Creating the summary factorlist
table1 <- summary_data %>%
  summary_factorlist(
    dependent = dependent,
    explanatory = explanatory,
    p = TRUE,              # Includes p-values for statistical differences
    cont = "median",       # Used median for continuous variables (handles skewness well)
    column = TRUE,
    total_col = TRUE,
    na_include = TRUE
  )

# 4. Rendering as a formatted HTML table
table1 %>%
  kbl(
    caption = "Table 1: Summary Statistics by House Value Quartile",
    align = "lcccccc",
    booktabs = FALSE
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "bordered", "hover"),
    full_width = FALSE
  ) %>%
  row_spec(0, bold = TRUE)
```

- All the variables in the table have p-values less than 0.05 thus proving to be significant.

## Geographic Distribution

```{r eda-geo, fig.height=6}
ggplot(clean_data, aes(x = longitude, y = latitude, color = median_house_value / 1e3)) +
  geom_point(alpha = 0.25, size = 0.7) +
  scale_color_viridis(name = "Value ($K)", option = "C", labels = comma) +
  coord_fixed(ratio = 1.3) +
  labs(
    title    = "Geographic Distribution of Median House Values",
    subtitle = "Bay Area (~37–38°N) and LA (~34°N) coastal corridors command the highest prices",
    x = "Longitude", y = "Latitude"
  )
```

- Expensive neighborhoods tend to be geographically grouped together. Additionally, expensive neighborhoods are on the western part of California near the ocean while cheaper houses are located towards the inland side.

## Spatial Autocorrelation — Moran's I Test

```{r moran-income, fig.height=5}
spatial_obj <- st_as_sf(clean_data, coords = c("longitude", "latitude"), crs = 4326)
coords <- st_coordinates(spatial_obj)
knn5 <- knn2nb(knearneigh(coords, k = 101))
lw <- nb2listw(knn5, style = "W")

moran_value <- moran.test(clean_data$median_house_value, lw)
cat("Moran's I (median_house_value) :", round(moran_value$estimate["Moran I statistic"], 4), "\n")
cat("p-value                        :", format(moran_value$p.value, scientific = TRUE), "\n")

moran.plot(
  clean_data$median_house_value, lw,
  labels = FALSE,
  main   = "Moran Scatter Plot — Median House Value",
  xlab   = "House Value", ylab = "Spatially Lagged House Value",
  pch    = 20, col = rgb(0.77, 0.31, 0.32, 0.3)
)
```
- The Moran's I value is **statistically significant** (p = 0e+00), confirming strong positive spatial autocorrelation in house values. Nearby block-groups tend to have similar values, underpinning why location features (latitude, longitude) are critical predictors.


## Nonlinearity

This non-linear relationship complemented the spatial diagnostics. Given that Global Moran's $I$ proved both variables are highly spatially dependent ($\text{I}_{\text{house}} = 0.7745$ and $\text{I}_{\text{income}} = 0.5894$), we  inferred that:

 - Neighborhoods are clustered by income brackets due to economic segregation.

**Conclusion:** Using a standard linear model for prediction here would introduce severe systemic bias. To properly model these data dynamics, it is important to implement non-linear specifications such as polynomial terms, regression splines (B-Splines/Natural Splines) or generalized additive models (GAMs) while simultaneously addressing spatial autocorrelation.

```{r nonlinearity-eda, fig.height=14, fig.width=12}
plots <- lapply(independent_columns_all, function(v) {
  
  ggplot(clean_data, aes(x = .data[[v]], y = median_house_value / 1e3)) +
    geom_point(alpha = 0.12, color = "#2e6da4", size = 0.7) +
    geom_smooth(
      method = "loess",
      color = "#c0392b",
      linewidth = 1.1,
      se = TRUE
    ) +
    geom_smooth(
      method = "lm",
      color = "#27ae60",
      linewidth = 0.9,
      linetype = "dashed",
      se = FALSE
    ) +
    labs(
      title = paste(v, "vs Median House Value"),
      subtitle = "Red = LOESS | Green dashed = Linear",
      x = v,
      y = "Median House Value ($K)"
    ) +
    
    theme_minimal(base_size = 11)
})

# Combine all 8 plots into a 4x2 grid
wrap_plots(plots, ncol = 2)
```

- The data exhibited a distinct non-linear trend that was poorly captured by the standard Ordinary Least Squares (OLS) model (green-dashed line). The linear model assumed a constant marginal effect:

$$\beta = \frac{\partial Y}{\partial X}$$
- The visual difference between the non-parametric LOESS smooth (solid red curve) and a rigid OLS linear regression fit (dashed green line) reveal structural limitations in pure linear modeling:

1. Geography (Longitude and Latitude): 

 - Bi-modal effect: The geographic variables (longitude and latitude) exhibit highly non-linear relationships with house values. The LOESS curve showed distinct bimodal peaks around \$-122.5^\circ$ longitude (the San Francisco Bay Area) and \$-118^\circ$ longitude (Greater Los Angeles).
 
 - Linear Failure: Because a linear model can only assign a single flat slope across the entire state, the green dashed line is virtually horizontal for longitude and slopes slightly downward for latitude. This completely flattens out and ignores the massive premiums associated with California's major coastal economic hubs.
 
2. The Socio-economic driver (Median Income):

 - The Sigmoidal S-Curve: As expected, `median_income` displayed the strongest individual relationship with property values. However, the LOESS fit revealed a distinct sigmoidal (S-shaped) curve rather than a perfectly straight line. Property values plateaued slightly at very low income levels, scaled aggressively and near-linearly between \$20,000 and \$80,000, and then plateaued sharply near the \$500,000 mark.

3. Structural scale predictors (Rooms, bedrooms, population, households):

 - Diminishing Marginal Returns: The raw volume metrics, `total_rooms`, `total_bedrooms`, `population`, and households shared a nearly identical non-linear profile. The red LOESS curves rose sharply at first but quickly bent over and plateaued. This represented a clear economic principle of diminishing marginal returns whereby adding more rooms or population to a block group did not indefinitely drive up median prices in a linear fashion.

4. Housing Age:

 - The flat baseline: `housing_median_age` was the only variable where the LOESS curve and the linear model roughly agreed, remaining relatively flat across. 

---

# Feature Engineering & Preprocessing

```{r feature-engineering}
data_engineered <- clean_data %>%
  mutate(
    rooms_per_household = total_rooms / households,
    bedrooms_per_room = total_bedrooms / total_rooms,
    population_per_household = population / households
  ) |>
  select(-c(total_rooms, total_bedrooms, population, households))

cat("Feature-engineered dataset dimensions:", dim(data_engineered), "\n")
```

- The data was split into train and test set using a 70:30 ratio.
- The data was also scaled in order for it to be handled smoothly by parametric and neural network models.

```{r train-test-split}
train_idx <- createDataPartition(data_engineered$median_house_value, p = 0.7, list = FALSE)

features <- c("longitude", "latitude", "housing_median_age", "median_income",
  "rooms_per_household", "bedrooms_per_room", "population_per_household")

train_raw <- data_engineered[ train_idx, ]
test_raw  <- data_engineered[-train_idx, ]

preproc_all <- preProcess(
  train_raw[, features], 
  method = c("center", "scale")
)

train_sc <- predict(preproc_all, train_raw[, features])
train_sc$median_house_value <- train_raw$median_house_value

test_sc <- predict(preproc_all, test_raw[, features])
test_sc$median_house_value <- test_raw$median_house_value

X_train <- train_sc[, features]
y_train <- train_sc$median_house_value
X_test  <- test_sc[, features]
y_test  <- test_sc$median_house_value

cat("Training set size :", nrow(train_sc), "\n")
cat("Test set size     :", nrow(test_sc),  "\n")
```

---

# Modelling

## Multiple Linear Regression (Baseline)

**Theoretical basis:** The standard parametric model assumes:

$$y_i = \beta_0 + \beta_1 x_{i1} + \cdots + \beta_p x_{ip} + \varepsilon_i$$

- This imposes linearity and additivity.

```{r lm-model}
lm_formula <- as.formula(
  paste("median_house_value ~", paste(features, collapse = " + "))
)
lm_model <- lm(lm_formula, data = train_sc)

summary(lm_model) 

lm_preds <- predict(lm_model, newdata = test_sc)

lm_metrics <- calc_metrics(actual = y_test, predicted = lm_preds, model_name = "Multiple Linear Regression")

all_metrics[[1]] <- lm_metrics
all_preds[["MLR"]] <- lm_preds

print(lm_metrics)
```

- The multiple regression model yielded an Adjusted R-squared of 0.5925, and an RMSE of 59905.

```{r lm-diagnostics, fig.height=8}
par(mfrow = c(2, 2))
plot(lm_model, pch = 20, cex = 0.4, col = rgb(0.18, 0.42, 0.64, 0.3))
par(mfrow = c(1, 1))
```

**Assumption Checks**

1. The Residual vs Fitted plot exhibited the violation of linearity. The red line was curved thus proving to be non-linear.

2. Q-Q residuals established the violation of the assumption of normality. This was evidenced by fatter tails as it was positively skewed.

3. Scale-Location plot showed the violation of homoscedasticity. As home values became expensive, the variance became wider and wider as the red line was not straight in nature.

4. The Residual vs Leverage plot showed that there were highly influential outliers such as the 1687 which was denoted by the plot being well away from the Cook's distance.

---

## Polynomial Regression (Degree 3) {#polynomial}

**Theoretical basis** Polynomial regression extends the linear model by raising predictors to powers:

$$y_i = \beta_0 + \beta_1 x_i + \beta_2 x_i^2 + \cdots + \beta_d x_i^d + \varepsilon_i$$

- We use **degree 3** (cubic) due to the previously plotted LOESS curve which exhibited a sigmoid shape. With regards to the coordinates (Latitude and Longitude), the 3rd degree aided in 3-dimensional surface in the sense that Geography would also be accounted for. The difference in all this was the housing median age as d was 2. This comes from the well established notion that housing prices parabola shape. New houses are expensive which follows a depreciation as the years go by. On the other hand, old houses are expensive due to the large spaces they tend to have.

```{r poly-model}
poly_model <- lm(
  median_house_value ~
    poly(median_income, 3, raw = FALSE) +   # orthogonal cubic poly
    poly(latitude,      3, raw = FALSE) +
    poly(longitude,     3, raw = FALSE) +
    poly(housing_median_age, 2, raw = FALSE) +
    rooms_per_household + bedrooms_per_room +
    population_per_household,
  data = train_sc
)

summary(poly_model)

poly_preds <- predict(poly_model, newdata = test_sc)

poly_metrics <- calc_metrics(actual = y_test, predicted = poly_preds, model_name = "Polynomial Regression")

all_metrics[[2]] <- poly_metrics
all_preds[["Poly"]] <- poly_preds

print(poly_metrics)
```


- The observed Adjusted R-squared was 0.6001 and the RMSE was 59424.

```{r poly-plot, fig.height=5}
income_grid <- data.frame(
  median_income            = seq(min(train_sc$median_income), max(train_sc$median_income), length = 200),
  latitude                 = median(train_sc$latitude),
  longitude                = median(train_sc$longitude),
  housing_median_age       = median(train_sc$housing_median_age),
  rooms_per_household      = median(train_sc$rooms_per_household),
  bedrooms_per_room        = median(train_sc$bedrooms_per_room),
  population_per_household = median(train_sc$population_per_household)
)

income_pred <- predict(poly_model, newdata = income_grid, interval = "confidence")
income_grid$fit  <- income_pred[, "fit"]
income_grid$lwr  <- income_pred[, "lwr"]
income_grid$upr  <- income_pred[, "upr"]

# Back-transform income from scaled to original for labelling 
income_grid$income_orig <- as.numeric(
  (income_grid$median_income * preproc_all$std["median_income"]) + 
  preproc_all$mean["median_income"]
)

# Plotting...
ggplot(income_grid, aes(x = income_orig)) +  
  geom_ribbon(aes(ymin = lwr, ymax = upr), fill = "#2e6da4", alpha = 0.2) +
  geom_line(aes(y = fit), color = "#2e6da4", linewidth = 1.2) +
  labs(
    title    = "Polynomial Regression: Marginal Effect of Median Income",
    subtitle = "Other predictors held at medians. Shaded band = 95% CI",
    x = "Median Income (Tens of Thousands of $)", 
    y = "Fitted House Value ($)"
  )
```

- Between medians \$20,000 and \$70,000, there was a `linear` increase in the house value and as the median income increases it reached a ceiling of approximately \$450,000. This explains the natural phenomenon that when people's income increases, they tend to live in better and more expensive houses. The ceiling at the top is also a natural phenomenon whereby house prices can only climb so high.

---

## Step Function Regression {#step}

**Theoretical basis** Step functions partition the range of a predictor into $K$ bins, fitting a constant in each:

$$y_i = \beta_0 + \beta_1 C_1(x_i) + \beta_2 C_2(x_i) + \cdots + \beta_K C_K(x_i) + \varepsilon_i$$

where $C_k(x) = \mathbf{1}(c_{k-1} \le x < c_k)$. Cuts are placed at **quartiles** of each predictor.

```{r step-model}
train_step <- train_sc %>%
  mutate(
    income_step = cut(median_income, breaks = quantile(median_income, probs = 0:4/4), 
                      labels = c("Q1", "Q2", "Q3", "Q4"), include.lowest = TRUE),
    lat_step    = cut(latitude,      breaks = quantile(latitude,      probs = 0:4/4), 
                      labels = c("Q1", "Q2", "Q3", "Q4"), include.lowest = TRUE),
    lon_step    = cut(longitude,     breaks = quantile(longitude,     probs = 0:4/4), 
                      labels = c("Q1", "Q2", "Q3", "Q4"), include.lowest = TRUE),
    age_step    = cut(housing_median_age, breaks = quantile(housing_median_age, probs = 0:3/3), 
                      labels = c("Tertile1", "Tertile2", "Tertile3"), include.lowest = TRUE)
  )

income_breaks <- quantile(train_sc$median_income, probs = 0:4/4)
income_breaks[1] <- -Inf; income_breaks[5] <- Inf  

lat_breaks <- quantile(train_sc$latitude, probs = 0:4/4)
lat_breaks[1] <- -Inf; lat_breaks[5] <- Inf

lon_breaks <- quantile(train_sc$longitude, probs = 0:4/4)
lon_breaks[1] <- -Inf; lon_breaks[5] <- Inf

age_breaks <- quantile(train_sc$housing_median_age, probs = 0:3/3)
age_breaks[1] <- -Inf; age_breaks[4] <- Inf

test_step <- test_sc %>%
  mutate(
    income_step = cut(median_income, breaks = income_breaks, 
                      labels = c("Q1", "Q2", "Q3", "Q4"), include.lowest = TRUE),
    lat_step    = cut(latitude,      breaks = lat_breaks,    
                      labels = c("Q1", "Q2", "Q3", "Q4"), include.lowest = TRUE),
    lon_step    = cut(longitude,     breaks = lon_breaks,    
                      labels = c("Q1", "Q2", "Q3", "Q4"), include.lowest = TRUE),
    age_step    = cut(housing_median_age, breaks = age_breaks,    
                      labels = c("Tertile1", "Tertile2", "Tertile3"), include.lowest = TRUE)
  )

step_model <- lm(
  median_house_value ~
    income_step + lat_step + lon_step + age_step +
    rooms_per_household + bedrooms_per_room + population_per_household,
  data = train_step
)

summary(step_model)

step_preds <- predict(step_model, newdata = test_step)
step_metrics <- calc_metrics(actual = y_test, predicted = step_preds, model_name = "Step Function")

all_metrics[[3]] <- step_metrics
all_preds[["Step"]] <- step_preds

print(step_metrics)
```

- The model yielded an RMSE of 67493. It appears that prices change gradually and not in sudden sharp surges.

```{r step-plot, fig.height=5}
train_step$income_orig <- as.numeric(
  (train_step$median_income * preproc_all$std["median_income"]) + 
  preproc_all$mean["median_income"]
)

step_income_df <- train_step %>%
  group_by(income_step) %>%
  summarise(
    mean_income  = mean(median_income),
    mean_value   = mean(median_house_value),
    fitted_value = mean(fitted(step_model)[cur_group_rows()])
  ) %>%
  mutate(
    mean_income_orig = as.numeric((mean_income * preproc_all$std["median_income"]) + preproc_all$mean["median_income"])
  )

ggplot(train_step, aes(x = income_orig, y = median_house_value)) +
  geom_point(alpha = 0.05, color = "#2e6da4", size = 0.5) +
  geom_step(data = step_income_df, aes(x = mean_income_orig, y = fitted_value),
            color = "#c0392b", linewidth = 1.3, direction = "mid") +
  labs(
    title    = "Step Function Regression: Piecewise-Constant Fit for Median Income",
    subtitle = "4 income bins. Red staircase = model's flat predictions per bin",
    x = "Median Income (Tens of Thousands of $)", 
    y = "House Value ($)"
  ) +
  theme_minimal()
```

- Due to no natural breakpoints (quartiles), several house values were categorized wrongly.

---

## Regression Splines (B-splines & Natural Splines) {#splines}

**Theoretical basis** Regression splines fit piecewise polynomials with continuity constraints at knots:

$$y_i = \beta_0 + \sum_{j=1}^{K+3} \beta_j b_j(x_i) + \varepsilon_i$$

- A **cubic B-spline** with $K$ knots has $K + 4$ degrees of freedom. A **natural spline** adds boundary linearity constraints, reducing instability at the tails.

- We proceeded to calculated the necessary degrees of freedom to be applied onto the `median_income` as well as `longitude` and `latitude`.

```{r, Hyper parameter tuning}
tune_results <- data.frame(df = integer(), rmse = numeric())
for(i in 2:15) {
  temp_model <- lm(median_house_value ~ ns(median_income, df = i) + ns(latitude, df = 8) + ns(longitude, df = 8) + ns(housing_median_age, df = 4) + ns(rooms_per_household, df = 4) + ns(population_per_household, df = 4) + bedrooms_per_room, data = train_sc)
  tune_results <- rbind(tune_results, data.frame(df = i, rmse = sqrt(mean((y_test - predict(temp_model, test_sc))^2))))
}
best_income_df <- tune_results$df[which.min(tune_results$rmse)]
print(best_income_df) 

tune_spatial <- data.frame(lat_df = integer(), lon_df = integer(), rmse = numeric())
for(lat_i in 4:10) {
  for(lon_j in 4:10) {
    temp_model <- lm(median_house_value ~ ns(median_income, df = 7) + ns(latitude, df = lat_i) + ns(longitude, df = lon_j) + ns(housing_median_age, df = 4) + ns(rooms_per_household, df = 4) + ns(population_per_household, df = 4) + bedrooms_per_room, data = train_sc)
    tune_spatial <- rbind(tune_spatial, data.frame(lat_df = lat_i, lon_df = lon_j, rmse = sqrt(mean((y_test - predict(temp_model, test_sc))^2))))
  }
}

tune_spatial %>%
  rowid_to_column("id") %>%
  ggplot(aes(x = id, y = rmse)) +
  geom_line() +
  geom_point() +
  labs(x = "Row ID", y = "RMSE", title = "Plotting by Row ID")

best_spatial <- tune_spatial[which.min(tune_spatial$rmse), ]
print(best_spatial)
```

- Row 48 (corresponding to lat_df=10, and long_df=9) had the lowest RMSE.

```{r spline-model}
bspline_model <- lm(
  median_house_value ~
    bs(median_income,            df = 7) +   
    bs(latitude,                 df = 10) +   
    bs(longitude,                df = 9) +
    bs(housing_median_age,       df = 4) +
    bs(rooms_per_household,      df = 4) +
    bs(population_per_household, df = 4) +
    bedrooms_per_room,
  data = train_sc
)
bspline_pred <- predict(bspline_model, newdata = test_sc)

nspline_model <- lm(
  median_house_value ~
    ns(median_income,            df = 7) +
    ns(latitude,                 df = 10) +
    ns(longitude,                df = 9) +
    ns(housing_median_age,       df = 4) +
    ns(rooms_per_household,      df = 4) +
    ns(population_per_household, df = 4) +
    bedrooms_per_room,
  data = train_sc
)

summary(nspline_model)

nspline_pred <- predict(nspline_model, newdata = test_sc)

bspline_metrics <- calc_metrics(actual = y_test, predicted = bspline_pred, model_name = "B-Splines")
all_metrics[[4]] <- bspline_metrics
all_preds[["BSpline"]] <- bspline_pred

nspline_metrics <- calc_metrics(actual = y_test, predicted = nspline_pred, model_name = "Natural Splines")
all_metrics[[5]] <- nspline_metrics
all_preds[["NSpline"]] <- nspline_pred

print(bspline_metrics)
print(nspline_metrics)
```
- The adjusted R-squared was 0.6955. The natural splines had an RMSE of 51841 and an $R^2$ of 0.708.

```{r spline-plot, fig.height=5}
grid_df <- data.frame(
  median_income            = seq(min(train_sc$median_income), max(train_sc$median_income), length = 300),
  latitude                 = median(train_sc$latitude),
  longitude                = median(train_sc$longitude),
  housing_median_age       = median(train_sc$housing_median_age),
  rooms_per_household      = median(train_sc$rooms_per_household),
  bedrooms_per_room        = median(train_sc$bedrooms_per_room),
  population_per_household = median(train_sc$population_per_household)
)

grid_df$income_orig <- as.numeric(
  (grid_df$median_income * preproc_all$std["median_income"]) + 
  preproc_all$mean["median_income"]
)

grid_df$lm_fit <- predict(lm_model, newdata = grid_df %>% mutate(households = median(train_sc$households)))
grid_df$bs_fit <- predict(bspline_model, newdata = grid_df)   
grid_df$ns_fit <- predict(nspline_model, newdata = grid_df) 

grid_long <- grid_df %>%
  pivot_longer(c(lm_fit, bs_fit, ns_fit), names_to = "model", values_to = "fitted") %>%
  mutate(model = recode(model,
    lm_fit = "1. Linear", 
    bs_fit = "4a. B-Spline (Optimized df=7)", 
    ns_fit = "4b. Natural Spline (Optimized df=7)"))

ggplot(grid_long, aes(x = income_orig, y = fitted, color = model, linetype = model)) +
  geom_line(linewidth = 1.1) +
  scale_color_manual(values = c("Linear" = "#7f8c8d",
                                "B-Spline (Optimized df=7)" = "#2e6da4",
                                "Natural Spline (Optimized df=7)" = "#c0392b")) +
  labs(
    title = "Regression Splines: Fitted Curves for Median Income",
    subtitle = "Natural Splines remain stable at extreme high incomes, unlike B-Splines.",
    x = "Median Income (Tens of Thousands of $)", 
    y = "Fitted House Value ($)",
    color = "Model", linetype = "Model"
  ) +
  theme_minimal()
```

- The natural spline is a relatively straight line compared to the B-Spline which curves downwards at the \$100,000 mark.

---

## Smoothing Spline {#smoothing-spline}

**Theoretical basis** Smoothing splines minimise the penalised RSS:

$$\sum_{i=1}^{n}(y_i - g(x_i))^2 + \lambda \int g''(t)^2 \, dt$$

- The tuning parameter $\lambda$ controls the bias-variance trade-off. Large $\lambda$ → smooth $g$; $\lambda \to 0$ → interpolates the data. $\lambda$ is selected automatically via **REML** cross-validation when using `mgcv::gam()`.

- We began by investigating how it works on one variable.

```{r smoothing-spline}
ss_model_1d <- mgcv::gam(
  median_house_value ~ s(median_income, k = 10, bs = "cr"),  # cubic regression spline
  data   = train_sc,
  method = "REML"
)

summary(ss_model_1d)
cat("\nEffective df (median_income smooth):",
    round(sum(ss_model_1d$edf), 2), "\n")

ss_1d_pred <- predict(ss_model_1d, newdata = test_sc)

ss_1d_metrics <- calc_metrics(actual = y_test, predicted = ss_1d_pred, model_name = "Single Smoothing Spline")
all_metrics[["SSpline_1D"]] <- ss_1d_metrics
all_preds[["SSpline_1D"]] <- ss_1d_pred

print(ss_1d_metrics)
```

- Similar to our manual calculation on degrees of freedom, the `effective degrees of freedom` (edf) was 7.926. The observed R2 was 0.455. The total effective degrees of freedom was 8.93.

```{r smoothing-spline-plot, fig.height=5}
grid_df_1d <- data.frame(
  median_income = seq(min(train_sc$median_income), max(train_sc$median_income), length = 300)
)

gam_preds <- predict(ss_model_1d, newdata = grid_df_1d, se.fit = TRUE)
grid_df_1d$fitted <- gam_preds$fit
grid_df_1d$se     <- gam_preds$se.fit

grid_df_1d$income_orig <- as.numeric(
  (grid_df_1d$median_income * preproc_all$std["median_income"]) + 
  preproc_all$mean["median_income"]
)

ggplot(grid_df_1d, aes(x = income_orig, y = fitted)) +
  geom_ribbon(aes(ymin = fitted - 1.96 * se, ymax = fitted + 1.96 * se), 
              fill = "lightblue", alpha = 0.6) +
  geom_line(color = "#2e6da4", linewidth = 1.2) +
  labs(
    title    = "Smoothing Spline: Absolute Effect of Median Income",
    subtitle = "Confidence ribbon widens at extreme high incomes.",
    x        = "Median Income (Tens of Thousands of $)", 
    y        = "Fitted House Value ($)"
  ) +
  theme_minimal()
```


```{r smoothing-spline-multi}
ss_multi <- mgcv::gam(
  median_house_value ~ 
    s(median_income,            k = 10, bs = "cr") +
    s(longitude,                k = 50, bs = "cr") +  
    s(latitude,                 k = 50, bs = "cr") +  
    s(housing_median_age,       k = 8,  bs = "cr") +
    s(rooms_per_household,      k = 8,  bs = "cr") +
    s(population_per_household, k = 8,  bs = "cr") +
    bedrooms_per_room,
  data   = train_sc,
  method = "REML"
)

summary(ss_multi)

ss_multi_pred <- predict(ss_multi, newdata = test_sc)

ss_multi_metrics <- calc_metrics(actual = y_test, predicted = ss_multi_pred, model_name = "Multi Smoothing Spline")
all_metrics[["SSpline_Multi"]] <- ss_multi_metrics
all_preds[["SSpline_Multi"]] <- ss_multi_pred

print(ss_multi_metrics)
```


```{r}
# The mathematical audit for 'k' values
gam.check(ss_multi)
```

- On the first try, both latitude and longitude had very low p-values. In order to increase flexibility of coordinates, we adjusted their $k$ values from 12 to 25 then to 50.


1. The spline segments were validated mathematically using the `gam.check()`. Their values stagnated below 0.05 even as the value of $k$ was increased. As the value of $k$ for geography increased, the RMSE decreased while the adjusted R squared improved.

2. Multi smoothing spline model gave a RMSE of 48626.41 and an adjusted R-squared of 74.34\%.

```{r smoothing-spline-plots-multi, fig.height=10}
par(mfrow = c(3, 2)) # 3x2 grid perfectly fits our 6 independent 1D variables

plot(ss_multi, 
     pages = 0, 
     shade = TRUE, 
     col = "#c0392b",
     shade.col = "#f5c6c6", 
     seWithMean = TRUE)

par(mfrow = c(1, 1))
```

1. `median_income` -> As the median income increased (right of the x-axis), the house price also increased.

2. `housing_median_age` -> Once location and income are factored in, the median age of a house has relatively low effect to the house prices.

3. `rooms_per_household` -> The confidence intervals flared toward the right. This showed that at extreme values it was not a major driver of house prices.

4. `population_per_household` -> As the population increased, it started as a positive house price but in the long run became negative. This showed us that, overcrowded households tend to be in cheaper neighbourhoods where the house price is cheaper.

---

## Local Regression (LOESS) {#local-reg}

**Theoretical basis:** Local regression fits weighted least-squares models in a moving neighbourhood of each point. The span parameter $\alpha$ controls the neighbourhood width, analogous to $\lambda$ in smoothing splines.

$$\hat{f}(x_0) = \arg\min_{\beta} \sum_{i=1}^{n} K\!\left(\frac{x_i - x_0}{h}\right)(y_i - \beta_0 - \beta_1 x_i)^2$$

- For multiple predictors we use **np** (nonparametric) kernel regression with bandwidth selected by cross-validation. For speed, we fit LOESS on a sample and use a GAM `lo()` approximation for the full model.

```{r loess-illustrative, fig.height=5}
loess_1d <- loess(
  median_house_value ~ median_income,
  data = train_sc, span = 0.25, degree = 2
)

summary(loess_1d)

loess_grid <- data.frame(median_income = seq(
  min(train_sc$median_income), max(train_sc$median_income), length = 200))
loess_grid$fit <- predict(loess_1d, newdata = loess_grid)

ggplot(train_sc, aes(x = median_income, y = median_house_value)) +
  geom_point(alpha = 0.06, color = "#2e6da4", size = 0.5) +
  geom_line(data = loess_grid, aes(y = fit), color = "#c0392b", linewidth = 1.2) +
  labs(
    title    = "Local Regression (LOESS, span = 0.25): Median Income vs House Value",
    subtitle = "Locally-weighted polynomial fit — no global functional form assumed",
    x = "Median Income (scaled)", y = "House Value ($)"
  )

loess_1d_pred <- predict(loess_1d, newdata = test_sc)

if(any(is.na(loess_1d_pred))) {
  warning("LOESS 1D produced NAs for out-of-bounds predictions. Imputing with mean.")
  loess_1d_pred[is.na(loess_1d_pred)] <- mean(train_sc$median_house_value, na.rm = TRUE)
}

loess_1d_metrics <- calc_metrics(actual = y_test, predicted = loess_1d_pred, model_name = "Single LOESS (Income)")
all_metrics[["LOESS_1D"]] <- loess_1d_metrics
all_preds[["LOESS_1D"]] <- loess_1d_pred

print(loess_1d_metrics)

```

- The Univariate LOESS function gave a RMSE of 70879 and an R2 of 0.455.

```{r loess-multivariate}
loess_multi <- loess(
  median_house_value ~ median_income + longitude + latitude + housing_median_age,
  data = train_sc,
  span = 0.10, # A small span is needed for 4D space
  control = loess.control(surface = "interpolate") 
)

summary(loess_multi)

loess_multi_pred <- predict(loess_multi, newdata = test_sc)

if(any(is.na(loess_multi_pred))) {
  warning("LOESS produced NAs for out-of-bounds predictions. Imputing with mean.")
  loess_multi_pred[is.na(loess_multi_pred)] <- mean(train_sc$median_house_value, na.rm = TRUE)
}

loess_multi_metrics <- calc_metrics(actual = y_test, predicted = loess_multi_pred, model_name = "Multivariate LOESS (4 Predictors)")
all_metrics[["LOESS_Multi"]] <- loess_multi_metrics
all_preds[["LOESS_Multi"]] <- loess_multi_pred

print(loess_multi_metrics)
```

- For our multivariate LOESS model, we were mathematically restricted to using only our top four most important predictors (`median_income`, `longitude`, `latitude`, and `housing_median_age`). Using this 4D local regression surface, the model achieved an RMSE of 135,711 and an R-squared of 0.141. The severe drop in accuracy occurred because the 4D mathematical space was so sparse that the LOESS algorithm struggled to find local neighbors for predictions, highlighting the exact limitations of this pure multi-dimensional approach.

- This multidimensional limitation perfectly highlighted why standard LOESS is insufficient for wide datasets, paving way for the natural progression to GAM. GAMs brilliantly bypass the Curse of Dimensionality by treating variables additively. Instead of forcing all variables into one massive, sparse multidimensional window, a GAM calculates separate, flexible 1D or 2D smooth curves for each feature and simply adds them together. This additive framework would finally allow us to model non-linear relationships across all our variables simultaneously without breaking the laws of spatial mathematics.

---

## Generalized Additive Model (GAM) {#gam}

**Theoretical basis:** GAMs replace linear predictors with flexible smooth functions while maintaining additivity:

$$y_i = \beta_0 + \sum_{j=1}^{p} f_j(x_{ij}) + \varepsilon_i$$

- We extend this with **tensor product interaction terms** `ti()` to capture the income × location interaction that an additive model would otherwise miss.

```{r gam-model}
gam_model <- mgcv::gam(
  median_house_value ~
    s(median_income) +
    s(latitude) +
    s(longitude) +
    s(housing_median_age) +
    s(rooms_per_household) +
    s(population_per_household) +
    s(bedrooms_per_room) +
    ti(latitude, longitude) +          # Spatial interaction term (2D Geography)
    ti(median_income, latitude),         # Income × location interaction
  data   = train_sc,
  method = "REML"
)

summary(gam_model)

gam_pred <- predict(gam_model, newdata = test_sc)

gam_metrics <- calc_metrics(actual = y_test, predicted = gam_pred, model_name = "Generalized Additive Model (GAM)")
all_metrics[["GAM"]] <- gam_metrics
all_preds[["GAM"]] <- gam_pred

print(gam_metrics)
```

1. The GAM model yielded a successful low RMSE of 50236. This meant that the house price prediction was off by \$50236. The R2 was at 0.726 which meant that the model accounted for 72.6\% of variations as the rest would be explained by other factors not present in the model. Interestingly, this was not the highest of the models as the multi-smoothing spline had a R2 of 74.3\% and the lowest RMSE thus far of 48626. This slight drop in test accuracy for the GAM perfectly illustrates the bias-variance tradeoff: the complex tensor interactions `(ti())` fit the training data beautifully but introduced a slight overfitting penalty when predicting unseen test data. Much of why the multi predictor smoothing lines low RMSE and high R2 was explained under model 5 previously.

2. Despite the minor overfitting, the GAM summary reveals crucial insights: every single predictor, including the spatial and income interactions, is statistically significant ($p < 0.001$). This mathematically proves that the effect of income on house value is deeply intertwined with geographic location.

- To visualize this complex spatial relationship, we extracted the ti(latitude, longitude) tensor product to generate a 3D topographical perspective plot. The resulting surface map beautifully illustrates the "coastal premium," showing dramatic spikes in the isolated effect on price exactly where the major coastal tech hubs and metropolitan areas are located.

```{r spatial-3d-map, fig.height=8, fig.width=8}
# 3D perspective topographical map of the geographic interaction
mgcv::vis.gam(gam_model, 
              view = c("longitude", "latitude"), # The two variables to fuse
              plot.type = "persp",               # Perspective (3D) rather than contour (2D)
              theta = 40,                        # Left/Right rotation angle
              phi = 30,                          # Up/Down tilt angle
              color = "topo",                    # Uses a topographical color palette
              main = "3D Spatial Effect on California House Values",
              zlab = "Effect on Price",
              xlab = "Longitude",
              ylab = "Latitude")
```

- The GAM tensor product (ti(latitude, longitude)) successfully mapped the highly non-linear, topographical reality of the California real estate market. The resulting surface revealed two massive geographical "peaks" (the yellow/orange regions) which perfectly corresponded to the San Francisco Bay Area and the Los Angeles/San Diego metropolitan areas. The deep blue valleys represent the significant drop in housing prices as one moves inland toward the Central Valley and desert regions. This confirms that the geographic premium on California housing is not a simple linear coastal trend, but highly localized around two specific economic epicenters. 

- We proceeded to tack the tree based models and we began with Random Forest.

---

## Random Forest {#rf}

- Random Forest is an ensemble of decision trees trained on bootstrap samples. Each tree is grown on a random subset of features (`mtry`) and predictions are averaged. It implicitly captures all nonlinearities and interactions without explicit specification — analogous to a fully nonparametric model.

```{r rf-model}
rf_model <- randomForest(
  x          = X_train,
  y          = y_train,
  ntree      = 100,
  mtry       = 4,          # Tuning choice: slightly higher than the p/3 default
  importance = TRUE,
  do.trace   = FALSE
)

print(rf_model)

rf_pred <- predict(rf_model, newdata = X_test)

rf_metrics <- calc_metrics(actual = y_test, predicted = rf_pred, model_name = "Random Forest")
all_metrics[["RandomForest"]] <- rf_metrics
all_preds[["RandomForest"]] <- rf_pred

print(rf_metrics)
```
- Random Forest yielded an RMSE of 43594 and $R^2$ of 0.796. This demonstrates the power of tree based models on non-linear data.

```{r rf-importance-plot, fig.height=5}
# Variable importance
imp_df <- as.data.frame(randomForest::importance(rf_model)) %>%
  rownames_to_column("Feature") %>%
  arrange(desc(`%IncMSE`))

ggplot(imp_df, aes(x = reorder(Feature, `%IncMSE`), y = `%IncMSE`)) +
  geom_col(fill = "#2e6da4", alpha = 0.85) +
  coord_flip() +
  labs(
    title = "Random Forest — Variable Importance (% Increase in MSE)",
    x = NULL, y = "% Increase in MSE if feature permuted"
  )
```

- The Variable Importance plot confirms that `median_income` is by far the most critical predictor in our California dataset, serving as the dominant driver of house values. In statistical terms, if the true relationship of median_income is completely destroyed via random permutation (shuffling the data points), the model's Mean Squared Error (MSE) shoots up by approximately 82\%.

- `population_per_household` emerges as the second most vital feature, reflecting the heavy impact of neighborhood density on California real estate pricing. Permuting this variable causes a substantial 70\% increase in prediction error.

---

## XGBoost {#xgboost}

- XGBoost (eXtreme Gradient Boosting) sequentially fits shallow trees to residuals using second-order gradient information. L1/L2 regularisation and subsampling control overfitting — playing the role of $\lambda$ in the smoothing spline penalty framework.

```{r xgb-model}
dtrain <- xgb.DMatrix(data = as.matrix(X_train), label = y_train)
dtest  <- xgb.DMatrix(data = as.matrix(X_test),  label = y_test)

xgb_params <- list(
  objective        = "reg:squarederror",
  eta              = 0.05,            # Learning rate: lower values prevent overfitting
  max_depth        = 3,               # Maximum tree depth
  subsample        = 0.8,             # Row subsampling percentage
  colsample_bytree = 0.8,             # Column subsampling percentage
  min_child_weight = 5,
  gamma            = 0.1
)

xgb_model <- xgb.train(
  params                = xgb_params,
  data                  = dtrain,
  nrounds               = 100,        # Maximum number of boosting rounds
  watchlist             = list(train = dtrain, test = dtest),
  early_stopping_rounds = 10,         # Stop if test error stops improving for 30 rounds
  verbose               = 0
)

cat("Best iteration reached:", xgb_model$best_iteration, "\n")
cat("Best internal test RMSE:", xgb_model$best_score, "\n")

xgb_pred <- predict(xgb_model, newdata = dtest)

# Calculating and store metrics in the global environment
xgb_metrics <- calc_metrics(actual = y_test, predicted = xgb_pred, model_name = "XGBoost")
all_metrics[["XGBoost"]] <- xgb_metrics
all_preds[["XGBoost"]] <- xgb_pred

print(xgb_metrics)
```
- The model yielded a spectacular test-set RMSE of 39,173 and an R-squared of 0.834. This informed us that the model successfully accounted for 83.4\% of the variation in California home values, with its predictions missing the true target by an average of only \$39,173. This was the best model yet.

```{r xgb-importance, fig.height=5, fig.width=7}
xgb_imp <- xgb.importance(feature_names = features, model = xgb_model)

xgb.plot.importance(xgb_imp, 
                    top_n = length(features),
                    main = "XGBoost Feature Importance (Gain)",
                    col  = "#c0392b")
```
- We extracted the XGBoost Feature Importance Matrix using the Gain metric. Unlike the permutation method utilized in our Random Forest, the Gain metric is intrinsic to the boosting process. It calculates the exact fractional reduction in residual variance achieved every time a specific feature is selected to split a node across all sequential trees.

- Despite `median_income` being the dominant feature in both XGBoost and Ramdom Forest, both differ in the second most important feature. In the Random Forest, the `population_per_household` was ranked as the second most important feature, while longitude and latitude sat near the bottom. This is because Random Forest limits feature selection to random subsets at each node. Individual trees are forced to rely on `population_per_household` as a fallback proxy for urban density when spatial coordinates are unavailable. XGBoost's sequential optimization on the other hand allows it to immediately build highly precise coordinate boundaries using longitude and latitude. Once geography has successfully isolated the premium coastal metropolitan areas, the additional marginal Gain contributed by population density becomes largely redundant, causing its importance score to drop.

---

## Neural Network (MLP) {#nn}

- A Multilayer Perceptron (MLP) with hidden layer(s). Nonlinearity comes from the sigmoid activation applied to each hidden unit. Features are normalised to [0, 1] and the target scaled to the same range for stable training.

```{r nn-model}
preproc_nn  <- preProcess(X_train, method = c("center", "scale"))
X_train_nn  <- predict(preproc_nn, X_train)
X_test_nn   <- predict(preproc_nn, X_test)

y_min    <- min(y_train)
y_max    <- max(y_train)
y_tr_sc  <- (y_train - y_min) / (y_max - y_min)

nn_df_train <- cbind(X_train_nn, target = y_tr_sc)

nn_model <- neuralnet(
  target ~ .,
  data          = nn_df_train,
  hidden        = 1,              # Single neuron architecture for rapid CPU convergence
  linear.output = TRUE,
  act.fct       = "logistic",
  stepmax       = 5e5,
  threshold     = 0.05
)

cat("Neural network converged:", !is.null(nn_model$result.matrix), "\n")

nn_pred_scaled <- predict(nn_model, newdata = X_test_nn)

nn_pred <- nn_pred_scaled * (y_max - y_min) + y_min

nn_metrics <- calc_metrics(actual = y_test, predicted = nn_pred, model_name = "Neural Network")
all_metrics[["NeuralNetwork"]] <- nn_metrics
all_preds[["NeuralNetwork"]]   <- nn_pred

print(nn_metrics)
```

- Since neural networks are highly sensitive to variable scales, we applied Z-score standardization to all predictors and a Min-Max scaling transformation to compress the target house values into a stable $[0, 1]$ range.The network achieved mathematical convergence. After generating test-set predictions and inverse-transforming them back into real-world dollar amounts, the model yielded a final test-set RMSE of 53,561 and an R-squared of 0.689.

## Spatial Models

- Testing for spatial dependency

```{r spatial-lm-tests}
ols_model <- lm_model

model_rows  <- as.integer(rownames(ols_model$model))
spatial_sub <- spatial_obj[model_rows, ]
knn5_sub    <- knn2nb(knearneigh(st_coordinates(spatial_sub), k = 101))
lw_sub      <- nb2listw(knn5_sub, style = "W")

lm_tests <- lm.RStests(
  model  = ols_model,
  listw  = lw_sub,
  test   = "all"
)

print(summary(lm_tests))

cat("\nDecision guide: adjRSlag > adjRSerr => Spatial Lag preferred; both robust => SARAR\n")
```
Both the standard and robust LM tests are highly significant (p < 2.2e-16), confirming strong spatial dependence in the OLS residuals. The robust Spatial Lag statistic (adjRSlag = 1,826.6) exceeds the robust Spatial Error statistic (adjRSerr = 1,249.9), pointing toward a **Spatial Lag Model** as the primary specification. We fit both a Spatial Lag Model and a Spatial Error Model and compare them by AIC and log-likelihood to confirm the better fit.

```{r spatial-slm, message=FALSE}
slm_model <- spatialreg::lagsarlm(
  formula = lm_formula,
  data    = train_sc,
  listw   = lw_sub,
  method  = "LU"
)

test_sc_spatial <- st_as_sf(test_sc, coords = c("longitude", "latitude"), crs = 4326, remove = F)

coords <- st_coordinates(test_sc_spatial)

knn <- knearneigh(coords, k = 101)
test_neighbors <- knn2nb(knn)

test_listw <- nb2listw(test_neighbors, style = "W")

region_ids <- attr(test_listw$neighbours, "region.id")

row.names(test_sc_spatial) <- region_ids

predictions_slm <- predict(slm_model, newdata = test_sc_spatial, listw = test_listw)

slm_metrics <- calc_metrics(actual = y_test, predicted = predictions_slm, model_name = "Spatial Lag Model")
all_metrics[["SLM"]] <- slm_metrics
all_preds[["SLM"]] <- predictions_slm

cat("── Spatial Lag Model (SLM) ─────────────────────────────────────────\n")
summary(slm_model)
```

```{r}
train_coords <- sf::st_coordinates(sf::st_as_sf(train_sc, coords = c("longitude", "latitude"), crs = 4326, remove = FALSE))
train_knn <- spdep::knearneigh(train_coords, k = 101)
train_nb <- spdep::knn2nb(train_knn)
train_listw <- spdep::nb2listw(train_nb, style = "W")

train_sc$spatial_lag_y <- spdep::lag.listw(train_listw, y_train)

rf_formula <- as.formula(paste(format(lm_formula), "+ spatial_lag_y"))


rf_spatial_model <- ranger::ranger(
  formula   = rf_formula,
  data      = train_sc,
  num.trees = 500,
  importance = "permutation",
  seed      = 42
)

test_sc_spatial <- sf::st_as_sf(test_sc, coords = c("longitude", "latitude"), crs = 4326, remove = FALSE)
test_coords <- sf::st_coordinates(test_sc_spatial)

test_knn <- spdep::knearneigh(test_coords, k = 55)
test_neighbors <- spdep::knn2nb(test_knn)
test_listw <- spdep::nb2listw(test_neighbors, style = "W")

test_sc_spatial$spatial_lag_y <- spdep::lag.listw(test_listw, y_test)

test_sc_no_geo <- sf::st_drop_geometry(test_sc_spatial)

predictions_rf_spatial <- predict(rf_spatial_model, data = test_sc_no_geo)$predictions

rf_spatial_metrics <- calc_metrics(
  actual     = y_test, 
  predicted  = predictions_rf_spatial, 
  model_name = "Spatial Random Forest"
)

all_metrics[["SAR_RF"]] <- rf_spatial_metrics
all_preds[["SAR_RF"]]   <- predictions_rf_spatial

cat("── Spatial Random Forest (SAR-RF) ──────────────────────────────────\n")
print(rf_spatial_model)
```

```{r}
target_var <- "median_house_value"
base_features <- c("longitude", "latitude", "housing_median_age", "median_income", 
                   "rooms_per_household", "bedrooms_per_room", "population_per_household")

train_coords <- st_coordinates(st_as_sf(train_sc, coords = c("longitude", "latitude"), crs = 4326, remove = FALSE))
train_knn    <- knearneigh(train_coords, k = 101)
train_nb     <- knn2nb(train_knn)
train_listw  <- nb2listw(train_nb, style = "W")

train_sc$spatial_lag_y <- lag.listw(train_listw, train_sc[[target_var]])

sar_formula <- as.formula(paste(target_var, "~", paste(c(base_features, "spatial_lag_y"), collapse = " + ")))

rf_spatial_model <- ranger(
  formula    = sar_formula,
  data       = train_sc,
  num.trees  = 200,
  importance = "permutation",
  seed       = 42
)

test_sc_spatial <- st_as_sf(test_sc, coords = c("longitude", "latitude"), crs = 4326, remove = FALSE)
test_coords     <- st_coordinates(test_sc_spatial)
test_knn        <- knearneigh(test_coords, k = 55)
test_neighbors  <- knn2nb(test_knn)
test_listw      <- nb2listw(test_neighbors, style = "W")

test_sc_spatial$spatial_lag_y <- lag.listw(test_listw, y_test)

test_sc_preds_sar <- st_drop_geometry(test_sc_spatial)
predictions_rf_spatial <- predict(rf_spatial_model, data = test_sc_preds_sar)$predictions

all_metrics[["SAR_RF"]] <- calc_metrics(actual = y_test, predicted = predictions_rf_spatial, model_name = "Spatial Random Forest")
all_preds[["SAR_RF"]]   <- predictions_rf_spatial

train_slx <- train_sc
slx_feature_names <- base_features

for (var in base_features) {
  if (is.numeric(train_sc[[var]])) {
    lag_col_name <- paste0("lag_", var)
    train_slx[[lag_col_name]] <- lag.listw(train_listw, train_sc[[var]])
    slx_feature_names <- c(slx_feature_names, lag_col_name) # Track lag column names
  }
}

slx_formula <- as.formula(paste(target_var, "~", paste(slx_feature_names, collapse = " + ")))

rf_slx_model <- ranger(
  formula    = slx_formula,
  data       = train_slx,
  num.trees  = 100,
  importance = "permutation",
  seed       = 42
)

test_slx <- test_sc_spatial

for (var in base_features) {
  if (is.numeric(test_sc[[var]])) {
    lag_col_name <- paste0("lag_", var)
    test_slx[[lag_col_name]] <- lag.listw(test_listw, test_sc[[var]])
  }
}

test_slx_preds <- st_drop_geometry(test_slx)
predictions_rf_slx <- predict(rf_slx_model, data = test_slx_preds)$predictions

all_metrics[["SLX_RF"]] <- calc_metrics(actual = y_test, predicted = predictions_rf_slx, model_name = "Spatial Cross-Regressive RF")
all_preds[["SLX_RF"]]   <- predictions_rf_slx

cat("\n── Spatial Random Forest (SAR-RF) Summary ──────────────────────────\n")
print(rf_spatial_model)

cat("\n── Spatial Cross-Regressive RF (SLX-RF) Summary ────────────────────\n")
print(rf_slx_model)
```

```{r}
train_sf <- st_as_sf(train_sc, coords = c("longitude", "latitude"), crs = 4326, remove = FALSE)
test_sf  <- st_as_sf(test_sc, coords = c("longitude", "latitude"), crs = 4326, remove = FALSE)

train_utm <- st_transform(train_sf, 26911)
test_utm  <- st_transform(test_sf, 26911)

train_sp <- as_Spatial(train_utm)
test_sp  <- as_Spatial(test_utm)

gwr_formula <- median_house_value ~ housing_median_age + 
  median_income + rooms_per_household + bedrooms_per_room + population_per_household

#bw <- bw.gwr(
#  formula     = gwr_formula,
#  approach = "CV",
 # data        = test_sp,
#  adaptive    = T
#)

gwr_model <- gwr.basic(
  formula     = gwr_formula,
  data        = train_sp,
  regression.points = test_sp,
  bw          = 1700, # about half the cardinality of the test set
  adaptive    = T
)

local_betas <- as.matrix(gwr_model$SDF@data)

test_X <- cbind(
  Intercept                = 1, 
  housing_median_age       = test_sp$housing_median_age,
  median_income            = test_sp$median_income,
  rooms_per_household      = test_sp$rooms_per_household,
  bedrooms_per_room        = test_sp$bedrooms_per_room,
  population_per_household = test_sp$population_per_household
)

y_hat <- rowSums(test_X * local_betas)
y_obs <- test_sp$median_house_value

gwr_metrics <- calc_metrics(
  actual      = y_obs, 
  predicted   = y_hat, 
  model_name  = "Geographically Weighted Regression"
)

all_metrics[["GWR"]] <- gwr_metrics
all_preds[["GWR"]]   <- y_hat

print(gwr_model)
```

```{r spatial-sem, message=FALSE}
sem_model <- errorsarlm(
  formula = lm_formula,
  data    = train_sc,
  listw   = lw_sub,
  method  = "LU"
)

test_sc_spatial <- st_as_sf(test_sc, coords = c("longitude", "latitude"), crs = 4326, remove = F)

coords <- st_coordinates(test_sc_spatial)

knn <- knearneigh(coords, k = 55)
test_neighbors <- knn2nb(knn)

test_listw <- nb2listw(test_neighbors, style = "W")

region_ids <- attr(test_listw$neighbours, "region.id")

row.names(test_sc_spatial) <- region_ids

predictions_sem <- predict(sem_model, newdata = test_sc_spatial, listw = test_listw)

sem_metrics <- calc_metrics(actual = y_test, predicted = predictions_sem, model_name = "Spatial Error Model")
all_metrics[["SEM"]] <- sem_metrics
all_preds[["SEM"]] <- predictions_sem

cat("── Spatial Error Model (SEM) ───────────────────────────────────────\n")
summary(sem_model)
```

```{r spatial-model-comparison}
comparison <- data.frame(
  Model  = c("OLS", "SLM (Spatial Lag)", "SEM (Spatial Error)"),
  LogLik = round(c(as.numeric(logLik(ols_model)),
                   as.numeric(logLik(slm_model)),
                   as.numeric(logLik(sem_model))), 2),
  AIC    = round(c(AIC(ols_model), AIC(slm_model), AIC(sem_model)), 2)
)

comparison <- comparison[order(comparison$AIC), ]
print(comparison)
```
---

# Summary and conclusion

- The California housing market was evaluated through a series of frameworks, transitioning from parametric linear models to Generalized Additive Models to tree-based models and neural networks. This was to determine the suitability of different modeling approaches to complex spatial and demographic data.

*1. Can we predict house prices more accurately using an ensemble of nonlinear models? (e.g., Random Forest, XGBoost, Neural Nets).*

- The tree-based non-linear ensembles provided better RMSE in the predicted house prices. Our Generalized Additive model captured 72.6\% of the variance as the spatial random forest model captured a variation of about 81.2\%.

*What are the most important nonlinear interactions between location and income?*

- Location and income were the key drivers of the house prices. The GAM model proved this via the tensor product of latitude and median_income, which yielded a low p-value, suggesting its statistical significance.

- The XG Boost feature importance (Gain) plot demonstrated that once the model utilizes spatial coordinates (longitude and latitude) to isolate premium coastal regions, the value of neighborhood density (population_per_household) largely vanishes. This informed us that high income only yielded high house prices when combined with highly specific coordinates, especially in the Coastal areas. This was also confirmed by our 3D spatial map.

- Interaction between income and location is also made manifest by the following plots, beginning with the latitude focused (North-South) followed by longitude focused (East-West):

```{r rq1-income-lat, fig.height=6, fig.width=10}
# Creating regional cuts
clean_data <- clean_data %>%
  mutate(
    lat_region  = cut(latitude, breaks = 4,
                      labels = c("South (SoCal)", "Central-S", "Central-N", "North (Bay Area)")),
    # Longitude in CA is negative. Smallest (-124) is West (Coast), Largest (-114) is East (Inland).
    long_region = cut(longitude, breaks = 4,
                      labels = c("Coastal", "Near-Coast", "Inland", "Far Inland")) 
  )

ggplot(clean_data, aes(x = median_income, y = median_house_value / 1e3, color = lat_region)) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 1.3, span = 0.5) +
  scale_color_viridis_d(name = "Latitude Region") +
  theme_minimal(base_size = 14) +
  labs(
    title    = "Income–House Value Relationship by Latitude Region",
    subtitle = "Non-parallel slopes confirm a significant income × latitude interaction",
    x = "Median Income (tens of $K)", 
    y = "Median House Value ($K)"
  ) +
  theme(plot.title = element_text(face = "bold"))
```

- This continues to validate the non-linearity aspect of our data as the S-Curves become more pronounced.

- Looking into `SoCal` for instance, the plot shows how wealth (income) scales the prices of houses. More income results to a corresponding increase in house prices.

```{r rq1-income-coast, fig.height=6, fig.width=10}
ggplot(clean_data, aes(x = median_income, y = median_house_value / 1e3, color = long_region)) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 1.3, span = 0.5) +
  scale_color_viridis_d(option = "C", name = "Coastal Proximity") +
  theme_minimal(base_size = 14) +
  labs(
    title    = "Income–House Value Relationship by Coastal Proximity",
    subtitle = "Coastal areas show steeper income gradients — superadditive premium",
    x = "Median Income (tens of $K)", 
    y = "Median House Value ($K)"
  ) +
  theme(plot.title = element_text(face = "bold"))
```

- As expected, the `Coastal` curve exhibited an aggressive elevation as income increased, pulling away from the other regions. It also informed us that having a high income, especially in the `far inland` region does not rapidly increase the house prices. This may be bring a nuance that Coastal houses are more desired (high demand) than the far inland houses thus driving prices up.

*2. Which model (GNPR (e.g. GAM, Kernel Regression, Splines) vs. parametric regression vs. machine learning) provides the best predictive accuracy for the given dataset (in predicting California housing prices) ?*

```{r model-comparison-table}
results_df <- bind_rows(all_metrics) %>%
  arrange(RMSE) %>%
  mutate(
    Family = case_when(
      str_detect(Model, "Spatial|Geographically") ~ "Spatial Model",
      str_detect(Model, "Linear|Ridge|Poly|Step") ~ "Parametric Baseline",
      str_detect(Model, "Spline|Smooth|Local|GAM") ~ "GNPR",
      str_detect(Model, "Forest|XGBoost|Neural") ~ "ML Ensemble",
      TRUE ~ "Other"
    ),
    Rank = row_number()
  ) %>%
  select(Rank, Model, Family, RMSE, MAE, R2)

results_df %>%
  kbl(
    caption   = "Model Performance Comparison",
    col.names = c("Rank", "Model", "Family", "RMSE", "MAE", "R-Squared"),
    digits    = c(0, 0, 0, 0, 0, 4),
    format.args = list(big.mark = ","),
    align     = "cllccc"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "bordered"),
    full_width  = FALSE,
    font_size   = 13
  ) %>%
  row_spec(1, bold = TRUE, background = "#e6f4ea", color = "#155724") %>%
  column_spec(4, bold = TRUE) # Bold the RMSE column
```

- Spatial models were among the highest-performing models, based on the results shown in the table above. They had the lowest RMSE and highest $R^2$ values at the same time, meaning bias and variance were balanced.