---
title: "California Housing Price Prediction: Beyond Linearity"
author: "Research Analysis"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 6)
```



# 1. Setup & Data Loading

```{r load-packages}
# Install missing packages if needed
pkgs <- c("tidyverse", "caret", "randomForest", "xgboost", "mgcv", "np",
          "splines", "corrplot", "gridExtra", "Metrics", "neuralnet",
          "ggplot2", "viridis", "pdp", "vip", "spdep", "sf", "kableExtra")

installed <- rownames(installed.packages())
to_install <- pkgs[!pkgs %in% installed]
if (length(to_install) > 0) install.packages(to_install, dependencies = TRUE)

library(tidyverse)
library(caret)
library(randomForest)
library(xgboost)
library(mgcv)
library(splines)
library(corrplot)
library(gridExtra)
library(Metrics)
library(neuralnet)
library(ggplot2)
library(viridis)
library(pdp)
library(vip)
library(finalfit)
library(spdep)
library(sf)
library(kableExtra)
```

Objectives:
1. To predict house prices using ensemble of non-linear, GNPR and Parametric regression models.
2. Identify the most important non-linear interactions between location and income
3. Compare the models from objective one above based on the predictive accuracy. 



```{r load-data}
data <- read.csv("D:/Module 3/Linear Models/California Housing Price project/cal_housing.data", header = FALSE) # Assuming no header row




# Assign column names

colnames(data) <- c("longitude", "latitude", "housing_median_age", 

                    "total_rooms", "total_bedrooms", "population", 

                    "households", "median_income", "median_house_value")

cat("Dimensions:", dim(data), "\n")
cat("Missing values:", sum(is.na(data)), "\n")
head(data)
```



# 2. Exploratory Data Analysis

```{r eda-summary}
summary(data)
```



```{r eda-distributions for numeric columns}
data_long <- data %>% 
  select(-c(longitude,latitude)) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value")

ggplot(data_long, aes(x = value)) +
  geom_histogram(bins = 40, fill = "steelblue", color = "white", alpha = 0.8) +
  #geom_vline(xintercept = median(x)) +
  facet_wrap(~variable, scales = "free") +
  theme_minimal() +
  labs(title = "Distribution of All Variables")
```

- The numeric features are skewed to the right apart from housing_median_age which is bi-modal.



```{r eda-correlation}
cor_matrix <- cor(data, use = "complete.obs")
corrplot(cor_matrix, method = "color", type = "upper",
         tl.cex = 0.8, addCoef.col = "black", number.cex = 0.6,
         title = "Correlation Matrix", mar = c(0,0,1,0))
```

- There is a very high correlation between total bedrooms, population, households and total rooms. Median income has a high correlation with median house value.

```{r eda-geo-map}
ggplot(data, aes(x = longitude, y = latitude, color = median_house_value)) +
  geom_point(alpha = 0.3, size = 0.8) +
  scale_color_viridis(name = "House Value ($)", labels = scales::comma) +
  theme_minimal() +
  labs(title = "Geographic Distribution of House Values",
       x = "Longitude", y = "Latitude")
```

-Expensive neighborhoods tend to be geographically grouped together. Additionally, expensive neighborhoods are on the western part of California near the ocean while cheaper houses are located towards the inland side.Due to urban planning, house prices seems to be regulated to a maximum of \$500001 and a minimum of approximately \$15000.



```{r}
ggplot(data, aes(x = longitude, y = latitude, color = median_income)) +
  geom_point(alpha = 0.3, size = 0.8) +
  scale_color_viridis(name = "Income Value ($)", labels = scales::comma) +
  theme_minimal() +
  labs(title = "Geographic Distribution of Income Values",
       x = "Longitude", y = "Latitude")
```



```{r eda income vs location}
spatial_object <- st_as_sf(data, coords = c("longitude", "latitude"), crs = 4326)

# create the nearest neighbour
coordinates <- st_coordinates(spatial_object)
knn <- knearneigh(coordinates, k =5)
neighbour <- knn2nb(knn)

# spartial weights
lw <- nb2listw(neighbour, style = "W")

# Moran's 1 test for median income
moran_test <- moran.test(data$median_income, lw)
print(moran_test)


# Moran Scatter plot
moran.plot(data$median_income, lw, labels = FALSE, main = "Spatial Correlation of Median Income")
```

- The correlation between location and median income is moderately positively correlated with a value of 0.5963337.



```{r}
spatial_object2 <- st_as_sf(data, coords = c("longitude", "latitude"), crs = 4326)

# create the nearest neighbour
coordinates <- st_coordinates(spatial_object2)
knn <- knearneigh(coordinates, k =5)
neighbour <- knn2nb(knn)

# spartial weights
lw <- nb2listw(neighbour, style = "W")

# Moran's 1 test for median income
moran_test <- moran.test(data$median_house_value, lw)
print(moran_test)


# Moran Scatter plot
moran.plot(data$median_house_value, lw, labels = FALSE, main = "Spatial Correlation of Median House Value")
```


- The correlation between location and median house value is highly positively correlated with a value of 0.8136747.




```{r Finalfit summary}
dependent <- "median_house_value"
explanatory <- data %>% 
  select(-c(median_house_value, longitude,latitude)) %>% names

data %>%
  finalfit::ff_plot(dependent, explanatory)
```


```{r}
# Create categorical dependent variable
data <- data %>%
  mutate(
    house_value_group = ntile(median_house_value, 4) %>%
      factor(labels = c("Low", "Medium", "High", "Very High"))
  )

dependent <- "house_value_group"

explanatory <- data %>%
  select(-median_house_value, -house_value_group, -longitude, -latitude) %>%
  names()

# Create finalfit table
table1 <- data %>%
  summary_factorlist(
    dependent = dependent,
    explanatory = explanatory,
    p = TRUE,
    cont = "median",
    column = TRUE,
    total_col = TRUE,
    na_include = TRUE
  )

# Render as formatted table
table1 %>%
  kbl(
    caption = "Summary Statistics by House Value Group",
    align = "lcccccc",
    booktabs = FALSE
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "bordered", "hover"),
    full_width = FALSE
  ) %>%
  row_spec(0, bold = TRUE)
```



# 3. Feature Engineering & Preprocessing


```{r preprocessing}
# Engineer derived features
data_clean <- data %>%
  mutate(
    rooms_per_household    = total_rooms / households,
    bedrooms_per_room      = total_bedrooms / total_rooms,
    population_per_household = population / households
  )

cat("Clean dataset dimensions:", dim(data_clean), "\n")
```

```{r scaling - standardization}
numeric <- data_clean %>%
  select(where(is.numeric)) %>%
  select(-c(longitude, latitude))

scaled_numeric <- numeric %>%
  scale(center = TRUE, scale = TRUE) %>%
  as.data.frame()
```



```{r}
head(scaled_numeric)
```


```{r train-test-split}
set.seed(42)
train_idx <- createDataPartition(scaled_numeric$median_house_value, p = 0.8, list = FALSE)
train_data <- scaled_numeric[train_idx, ]
test_data  <- scaled_numeric[-train_idx, ]

# Feature matrix (exclude target and log-target)
features <- c("housing_median_age", "median_income",
              "rooms_per_household", "bedrooms_per_room", "population_per_household")

X_train <- train_data[, features]
y_train <- train_data$median_house_value
X_test  <- test_data[, features]
y_test  <- test_data$median_house_value

cat("Train size:", nrow(X_train), "| Test size:", nrow(X_test), "\n")
```




# 4. Model 1 — Parametric Linear Regression (Baseline)

```{r linear-regression}
lm_formula <- as.formula(paste("median_house_value ~", paste(features, collapse = " + ")))
lm_model <- lm(lm_formula, data = train_data)
summary(lm_model)

lm_pred  <- predict(lm_model, newdata = test_data)
lm_rmse  <- rmse(y_test, lm_pred)
lm_r2    <- cor(y_test, lm_pred)^2

cat("\nLinear Regression — RMSE:", round(lm_rmse, 2), "| R²:", round(lm_r2, 4), "\n")
```


```{r income vs house value}
ggplot(data, aes(x = median_income, y = median_house_value)) +
  geom_point(alpha = 0.2, color = "steelblue") +
  geom_smooth(method = "loess", color = "red", se = TRUE) +
  theme_minimal() +
  labs(title = "Median Income vs House Value (with Locally Estimated Scatter plot Smoothing)",
       x = "Median Income", y = "Median House Value ($)")
```




```{r median age vs house value}
ggplot(data_clean, aes(x = housing_median_age, y = median_house_value)) +
  geom_point(alpha = 0.2, color = "steelblue") +
  geom_smooth(method = "loess", color = "red", se = TRUE) +
  theme_minimal() +
  labs(title = "Median Income vs House median Age (with Locally Estimated Scatter plot Smoothing)",
       x = "House median Age", y = "Median House Value ($)")
```



```{r}
ggplot(data_clean, aes(x = rooms_per_household, y = median_house_value)) +
  geom_point(alpha = 0.2, color = "steelblue") +
  geom_smooth(method = "loess", color = "red", se = TRUE) +
  theme_minimal() +
  labs(title = "Rooms per Household vs House Value (with Locally Estimated Scatter plot Smoothing)",
       x = "Rooms per Household", y = "Median House Value ($)")
```



```{r}
ggplot(data_clean, aes(x = bedrooms_per_room, y = median_house_value)) +
  geom_point(alpha = 0.2, color = "steelblue") +
  geom_smooth(method = "loess", color = "red", se = TRUE) +
  theme_minimal() +
  labs(title = "Median Income vs Bedroom per Room (with Locally Estimated Scatter plot Smoothing)",
       x = "Bedroom per Room", y = "Median House Value ($)")
```


```{r}
ggplot(data_clean, aes(x = population_per_household, y = median_house_value)) +
  geom_point(alpha = 0.2, color = "steelblue") +
  geom_smooth(method = "loess", color = "red", se = TRUE) +
  theme_minimal() +
  labs(title = "Median Income vs Population per Household (with Locally Estimated Scatter plot Smoothing)",
       x = "Population per Household", y = "Median House Value ($)")
```







# GNPR models
# 5. Model 2 — Polynomial Regression
$$y_i = \beta_0 + \beta_1 x_i + \beta_2 x_i^2 + \cdots + \beta_d x_i^d + \varepsilon_i$$


```{r Polynomial Regression}
# ── 1. Find optimal degree for each predictor via BIC ────────────────────────
predictors <- c("housing_median_age", "median_income", 
                "rooms_per_household", "bedrooms_per_room", 
                "population_per_household")

max_degree <- 5   # maximum degree to test

best_degrees <- sapply(predictors, function(var) {
  
  bic_scores <- sapply(1:max_degree, function(deg) {
    formula <- as.formula(
      paste("median_house_value ~ poly(", var, ",", deg, ")")
    )
    m <- lm(formula, data = train_data)
    BIC(m)
  })
  
  cat(var, "→ BIC by degree:", round(bic_scores, 1), 
      "→ Best degree:", which.min(bic_scores), "\n")
  
  which.min(bic_scores)   # return degree with lowest BIC
})

cat("\nSelected degrees:\n")
print(best_degrees)

# ── 2. Build final model using BIC-selected degrees ──────────────────────────
poly_terms <- paste0(
  "poly(", names(best_degrees), ", ", best_degrees, ")",
  collapse = " + "
)

final_formula <- as.formula(
  paste("median_house_value ~", poly_terms)
)

cat("\nFinal formula:\n")
print(final_formula)

poly_model <- lm(final_formula, data = train_data)

# ── 3. Evaluate on test set ───────────────────────────────────────────────────
poly_pred <- predict(poly_model, newdata = test_data)
poly_rmse <- rmse(y_test, poly_pred)
poly_r2   <- cor(y_test, poly_pred)^2

cat("\nPolynomial Regression (BIC-optimised degrees)\n")
cat("RMSE:", round(poly_rmse, 2), "| R²:", round(poly_r2, 4), "\n")
```

```{r}

```





# 6. Model 3 - Regression splines
$$y_i = \beta_0 + \sum_{j=1}^{K+3} \beta_j b_j(x_i) + \varepsilon_i$$


```{r}
# identify the degree of freedom
model <- gam(median_house_value ~ s(median_income, k = 10) +
    s(housing_median_age, k = 10) +
    s(rooms_per_household, k = 10) +
    s(population_per_household, k = 10) +
    s(bedrooms_per_room, k =10),
  data = train_data, method = "REML")

summary(model)
```



```{r Regression splines}
spline_model <- lm(
  median_house_value ~
    bs(median_income, df = 8) +
    bs(housing_median_age, df = 7) +
    bs(rooms_per_household, df = 8) +
    bs(population_per_household, df = 9) +
    bs(bedrooms_per_room, df = 7),
  data = train_data
)

spline_pred <- predict(spline_model, newdata = test_data)
spline_rmse <- rmse(y_test, spline_pred)
spline_r2   <- cor(y_test, spline_pred)^2

cat("Spline Regression — RMSE:", round(spline_rmse, 2), "| R²:", round(spline_r2, 4), "\n")
```


# 7. Model 4 - Smoothing splines
$$\sum_{i=1}^{n}(y_i - g(x_i))^2 + \lambda \int g''(t)^2 \, dt$$
```{r Smoothing splines}
# Single-predictor smoothing spline for median_income (illustrative)
ss_model_1d <- gam(
  median_house_value ~ s(median_income, k = 10, bs = "cr"),  # cubic regression spline
  data   = train_sc,
  method = "REML"
)

summary(ss_model_1d)
cat("\nEffective df (median_income smooth):",
    round(sum(ss_model_1d$edf), 2), "\n")
```



```{r}
# Plot the smoothing spline with uncertainty band
plot(ss_model_1d, shade = TRUE, col = "#2e6da4", shade.col = "lightblue",
     seWithMean = TRUE, residuals = TRUE, pch = 20, cex = 0.1,
     main = "Smoothing Spline: s(median_income) — REML-selected λ",
     xlab = "Median Income (scaled)", ylab = "Effect on House Value ($)")
abline(h = 0, lty = 2, col = "grey50")
```



```{r}
# Multi-predictor smoothing spline model
ss_multi <- gam(
  median_house_value ~
    s(median_income,            k = 10, bs = "cr") +
    s(latitude,                 k = 12, bs = "cr") +
    s(longitude,                k = 12, bs = "cr") +
    s(housing_median_age,       k =  8, bs = "cr") +
    s(rooms_per_household,      k =  8, bs = "cr") +
    s(population_per_household, k =  8, bs = "cr") +
    bedrooms_per_room,
  data   = train_sc,
  method = "REML"
)

ss_pred <- predict(ss_multi, newdata = test_sc)

all_preds[["SmoothingSpline"]] <- ss_pred
all_metrics[["SmoothingSpline"]] <- calc_metrics(y_test, ss_pred, "5. Smoothing Spline (REML)")

cat("Smoothing Spline — RMSE:", all_metrics$SmoothingSpline$RMSE,
    "| R²:", all_metrics$SmoothingSpline$R2, "\n")
```



```{r}
par(mfrow = c(3, 2))
plot(ss_multi, pages = 0, shade = TRUE, col = "#c0392b",
     shade.col = "#f5c6c6", seWithMean = TRUE,
     main = "Smoothing Spline Terms")
par(mfrow = c(1, 1))
```




# 8. Model 5 - Step function
```{r Step function}
# Create step (piecewise-constant) predictors using quartile cut-points
train_step <- train_data %>%
  mutate(
    income_step = cut(median_income, breaks = quantile(median_income, probs = 0:4/4),
                      include.lowest = TRUE),
    age_step    = cut(housing_median_age,      breaks = quantile(housing_median_age,      probs = 0:4/4),
                      include.lowest = TRUE),
    room_step    = cut(rooms_per_household,     breaks = quantile(rooms_per_household,     probs = 0:4/4),
                      include.lowest = TRUE),
    population_step    = cut(population_per_household, breaks = quantile(population_per_household, probs = 0:3/3),
                      include.lowest = TRUE),
    bedroom_step    = cut(bedrooms_per_room, breaks = quantile(bedrooms_per_room, probs = 0:3/3),
                      include.lowest = TRUE)
  )

# Save cut-points for test set
income_breaks <- quantile(train_data$median_income,      probs = 0:4/4)
age_breaks    <- quantile(train_data$housing_median_age,            probs = 0:4/4)
room_breaks    <- quantile(train_data$rooms_per_household,           probs = 0:4/4)
population_breaks    <- quantile(train_data$population_per_household,  probs = 0:3/3)
bedroom_breaks     <- quantile(train_data$bedrooms_per_room,  probs = 0:3/3)

test_step <- test_data %>%
  mutate(
    income_step = cut(median_income,      breaks = income_breaks, include.lowest = TRUE),
    age_step    = cut(housing_median_age, breaks = age_breaks,    include.lowest = TRUE),
    room_step    = cut(rooms_per_household, breaks = room_breaks,    include.lowest = TRUE),
    population_step    = cut(population_per_household, breaks = population_breaks,    include.lowest = TRUE),
    bedroom_step    = cut(bedrooms_per_room, breaks = bedroom_breaks,    include.lowest = TRUE)
  )

step_model <- lm(
  median_house_value ~
    income_step + age_step +
     room_step + population_step + bedroom_step,
  data = train_step
)

summary(step_model)
```

```{r}
# Visualise step fit for median income
step_income_df <- train_step %>%
  group_by(income_step) %>%
  summarise(
    mean_income  = mean(median_income),
    mean_value   = mean(median_house_value),
    fitted_value = mean(fitted(step_model)[cur_group_rows()])
  )

ggplot(train_step, aes(x = median_income, y = median_house_value)) +
  geom_point(alpha = 0.05, color = "#2e6da4", size = 0.5) +
  geom_step(data = step_income_df, aes(x = mean_income, y = fitted_value),
            color = "#c0392b", linewidth = 1.3, direction = "mid") +
  labs(
    title    = "Step Function Regression: Piecewise-Constant Fit for Median Income",
    subtitle = "4 income bins (quartile cut-points). Red step line = model fitted values per bin",
    x = "Median Income (scaled)", y = "House Value ($)"
  )
```




```{r}
step_pred <- predict(step_model, newdata = test_step)

all_preds[["StepFunction"]] <- step_pred
all_metrics[["StepFunction"]] <- calc_metrics(y_test, step_pred, "3. Step Function")

cat("Step Function — RMSE:", all_metrics$StepFunction$RMSE,
    "| R²:", all_metrics$StepFunction$R2, "\n")
```



# 9. Model 6 - Local Regression
**Theoretical basis:** Local regression fits weighted least-squares models in a
moving neighbourhood of each point. The span parameter $\alpha$ controls the
neighbourhood width — analogous to $\lambda$ in smoothing splines.

$$\hat{f}(x_0) = \arg\min_{\beta} \sum_{i=1}^{n} K\!\left(\frac{x_i - x_0}{h}\right)(y_i - \beta_0 - \beta_1 x_i)^2$$


For multiple predictors we use **np** (nonparametric) kernel regression with
bandwidth selected by cross-validation. For speed, we fit LOESS on a sample and
use a GAM `lo()` approximation for the full model.
```{r Local Regression}
# ── Illustrative 1D LOESS: income vs house value ──────────────────────────────
loess_1d <- loess(
  median_house_value ~ median_income,
  data = train_data, span = 0.25, degree = 2
)

loess_grid <- data.frame(median_income = seq(
  min(train_data$median_income), max(train_data$median_income), length = 200))
loess_grid$fit <- predict(loess_1d, newdata = loess_grid)

ggplot(train_data, aes(x = median_income, y = median_house_value)) +
  geom_point(alpha = 0.06, color = "#2e6da4", size = 0.5) +
  geom_line(data = loess_grid, aes(y = fit), color = "#c0392b", linewidth = 1.2) +
  labs(
    title    = "Local Regression (LOESS, span = 0.25): Median Income vs House Value",
    subtitle = "Locally-weighted polynomial fit — no global functional form assumed",
    x = "Median Income (scaled)", y = "House Value ($)"
  )
```



```{r}
# ── Multivariate local regression via GAM lo() ────────────────────────────────
# mgcv::gam with lo() terms provides a practical multivariate LOESS approximation
loess_model <- gam(
  median_house_value ~
    gam::lo(median_income, span = 0.3) +
    gam::lo(housing_median_age, span = 0.4) +
    gam::lo(rooms_per_household, span = 0.4) +
    rooms_per_household + bedrooms_per_room +
    population_per_household,
  data = train_data
)

loess_pred <- predict(loess_model, newdata = test_sc)

all_preds[["LocalRegression"]] <- loess_pred
all_metrics[["LocalRegression"]] <- calc_metrics(y_test, loess_pred, "6. Local Regression (LOESS)")

cat("Local Regression — RMSE:", all_metrics$LocalRegression$RMSE,
    "| R²:", all_metrics$LocalRegression$R2, "\n")
```




# 10. Model 7 — Generalized Additive Model (GAM)


**Theoretical basis (DSA 8305, §22–23):** GAMs replace linear predictors with
flexible smooth functions while maintaining additivity:

$$y_i = \beta_0 + \sum_{j=1}^{p} f_j(x_{ij}) + \varepsilon_i$$


We extend this with **tensor product interaction terms** `ti()` to capture the
income × location interaction that an additive model would otherwise miss.
```{r gam-model}
gam_model <- gam(
  median_house_value ~
    s(median_income, k = 10) +
    s(housing_median_age, k = 8) +
    s(rooms_per_household, k = 8) +
    s(population_per_household, k = 8),
  data = train_data,
  method = "REML"
)

summary(gam_model)
```

```{r gam-plots}
par(mfrow = c(3, 3))
plot(
  gam_model, pages = 0,
  shade     = TRUE,
  col       = "#2e6da4",
  shade.col = "lightblue",
  seWithMean = TRUE,
  residuals  = TRUE,
  pch = 20, cex = 0.15
)
par(mfrow = c(1, 1))
```


```{r}
# ── GAM: income × latitude interaction surface ──────────────────────────────
vis.gam(
  gam_model,
  view      = c("median_income", "latitude"),
  plot.type = "contour",
  color     = "terrain",
  too.far   = 0.05,
  main      = "GAM Tensor Interaction: Median Income × Latitude",
  xlab      = "Median Income (scaled)",
  ylab      = "Latitude (scaled)"
)
```


```{r}
vis.gam(
  gam_model,
  view      = c("longitude", "latitude"),
  plot.type = "contour",
  color     = "topo",
  too.far   = 0.05,
  main      = "GAM Spatial Interaction: Longitude × Latitude",
  xlab      = "Longitude (scaled)",
  ylab      = "Latitude (scaled)"
)
```



```{r}
gam_pred <- predict(gam_model, newdata = test_sc)

all_preds[["GAM"]] <- gam_pred
all_metrics[["GAM"]] <- calc_metrics(y_test, gam_pred, "7. GAM (+ tensor interactions)")

cat("GAM — RMSE:", all_metrics$GAM$RMSE, "| R²:", all_metrics$GAM$R2, "\n")
```




# Ensemble models
# 7. Model 8 — Random Forest

```{r random-forest}
set.seed(42)
rf_model <- randomForest(
  x = X_train, y = y_train,
  ntree = 300, mtry = 4,
  importance = TRUE
)

rf_pred <- predict(rf_model, newdata = X_test)
rf_rmse <- rmse(y_test, rf_pred)
rf_r2   <- cor(y_test, rf_pred)^2

cat("Random Forest — RMSE:", round(rf_rmse, 2), "| R²:", round(rf_r2, 4), "\n")
```




```{r rf-importance}
importance_df <- as.data.frame(importance(rf_model)) %>%
  rownames_to_column("Feature") %>%
  arrange(desc(`%IncMSE`))

ggplot(importance_df, aes(x = reorder(Feature, `%IncMSE`), y = `%IncMSE`)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Random Forest — Variable Importance (%IncMSE)",
       x = "Feature", y = "% Increase in MSE")
```




# 8. Model 9 — XGBoost

```{r xgboost}
dtrain <- xgb.DMatrix(data = as.matrix(X_train), label = y_train)
dtest  <- xgb.DMatrix(data = as.matrix(X_test),  label = y_test)

xgb_params <- list(
  objective        = "reg:squarederror",
  eta              = 0.05,
  max_depth        = 6,
  subsample        = 0.8,
  colsample_bytree = 0.8,
  min_child_weight = 5
)

set.seed(42)
xgb_model <- xgb.train(
  params  = xgb_params,
  data    = dtrain,
  nrounds = 500,
  watchlist = list(train = dtrain, test = dtest),
  early_stopping_rounds = 30,
  verbose = 0
)

xgb_pred <- predict(xgb_model, dtest)
xgb_rmse <- rmse(y_test, xgb_pred)
xgb_r2   <- cor(y_test, xgb_pred)^2

cat("XGBoost — RMSE:", round(xgb_rmse, 2), "| R²:", round(xgb_r2, 4), "\n")
```




```{r xgb-importance}
xgb_imp <- xgb.importance(feature_names = features, model = xgb_model)
xgb.plot.importance(xgb_imp, top_n = 11,
                    main = "XGBoost Feature Importance")
```




# 9. Model 10 — Neural Network

```{r neural-net}
# Scale features for neural network
preproc  <- preProcess(X_train, method = c("center", "scale"))
X_train_scaled <- predict(preproc, X_train)
X_test_scaled  <- predict(preproc, X_test)

# Scale target to [0,1]
y_min <- min(y_train); y_max <- max(y_train)
y_train_scaled <- (y_train - y_min) / (y_max - y_min)

nn_data_train <- cbind(X_train_scaled, target = y_train_scaled)

set.seed(42)
nn_model <- neuralnet(
  target ~ .,
  data       = nn_data_train,
  hidden     = 1,
  linear.output = TRUE,
  stepmax    = 1e6,
  threshold  = 0.05
)

nn_pred_scaled <- predict(nn_model, X_test_scaled)[, 1]
nn_pred        <- nn_pred_scaled * (y_max - y_min) + y_min
nn_rmse        <- rmse(y_test, nn_pred)
nn_r2          <- cor(y_test, nn_pred)^2

cat("Neural Network — RMSE:", round(nn_rmse, 2), "| R²:", round(nn_r2, 4), "\n")
```




# 10. Research Question 1: Nonlinear Interactions — Location × Income

```{r interaction-analysis}
# Visualise income effect across geographic regions
data_clean <- data_clean %>%
  mutate(
    lat_group  = cut(latitude,  breaks = 4, labels = c("South", "Central-S", "Central-N", "North")),
    long_group = cut(longitude, breaks = 4, labels = c("Coast", "Near-Coast", "Inland", "Far-Inland"))
  )

ggplot(data_clean, aes(x = median_income, y = median_house_value, color = lat_group)) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 1.2) +
  scale_color_viridis_d(name = "Latitude Region") +
  theme_minimal() +
  labs(title = "Income–House Value Relationship by Latitude Region",
       subtitle = "Nonlinear interactions: income effect varies by location",
       x = "Median Income", y = "Median House Value ($)")
```




```{r pdp-income-latitude}
# Partial Dependence Plot: income × latitude interaction (Random Forest)
pd_income <- partial(rf_model, pred.var = "median_income",
                     train = X_train, grid.resolution = 30)
pd_lat    <- partial(rf_model, pred.var = "latitude",
                     train = X_train, grid.resolution = 30)

p1 <- autoplot(pd_income, smooth = TRUE) +
  theme_minimal() + labs(title = "PDP: Median Income")

p2 <- autoplot(pd_lat, smooth = TRUE) +
  theme_minimal() + labs(title = "PDP: Latitude")

grid.arrange(p1, p2, ncol = 2)
```




```{r 2d-pdp}
# 2D PDP: income × latitude
pd_2d <- partial(rf_model,
                 pred.var = c("median_income", "latitude"),
                 train = X_train, grid.resolution = 20)

plotPartial(pd_2d, levelplot = TRUE, zlab = "House Value",
            colorkey = TRUE, col.regions = viridis(100),
            main = "2D PDP: Income × Latitude Interaction (Random Forest)")
```



```{r gam-interaction-plot}
# GAM tensor product smooth: income × latitude
vis.gam(gam_model, view = c("median_income", "latitude"),
        plot.type = "contour", color = "terrain",
        main = "GAM: Income × Latitude Interaction Surface",
        too.far = 0.05)
```




# 11. Research Question 2: Model Comparison

```{r model-comparison}
results <- tibble(
  Model = c("Linear Regression", "Spline Regression", "GAM",
            "Random Forest", "XGBoost", "Neural Network"),
  RMSE  = c(lm_rmse, spline_rmse, gam_rmse, rf_rmse, xgb_rmse, nn_rmse),
  R2    = c(lm_r2,   spline_r2,   gam_r2,   rf_r2,   xgb_r2,   nn_r2)
) %>%
  arrange(RMSE)

print(results)
```



```{r comparison-plot}
results_long <- results %>%
  pivot_longer(c(RMSE, R2), names_to = "Metric", values_to = "Value")

ggplot(results_long, aes(x = reorder(Model, Value), y = Value, fill = Model)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~Metric, scales = "free_y") +
  coord_flip() +
  scale_fill_viridis_d() +
  theme_minimal() +
  labs(title = "Model Comparison: RMSE and R²",
       x = "Model", y = "Value")
```




```{r residual-comparison}
residuals_df <- tibble(
  Actual     = y_test,
  LM         = lm_pred,
  Spline     = spline_pred,
  GAM        = gam_pred,
  RF         = rf_pred,
  XGBoost    = xgb_pred,
  NeuralNet  = nn_pred
) %>%
  pivot_longer(-Actual, names_to = "Model", values_to = "Predicted") %>%
  mutate(Residual = Actual - Predicted)

ggplot(residuals_df, aes(x = Predicted, y = Residual, color = Model)) +
  geom_point(alpha = 0.15, size = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  facet_wrap(~Model, scales = "free") +
  scale_color_viridis_d() +
  theme_minimal() +
  labs(title = "Residual Plots by Model",
       x = "Predicted Value", y = "Residual")
```



```{r actual-vs-predicted}
ggplot(residuals_df, aes(x = Actual, y = Predicted, color = Model)) +
  geom_point(alpha = 0.15, size = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "black", linetype = "dashed") +
  facet_wrap(~Model) +
  scale_color_viridis_d() +
  theme_minimal() +
  labs(title = "Actual vs Predicted by Model",
       x = "Actual House Value ($)", y = "Predicted House Value ($)")
```




# 12. Summary & Conclusions

```{r final-summary}
cat("=== FINAL MODEL PERFORMANCE SUMMARY ===\n\n")
print(results)

best_model <- results$Model[1]
cat("\nBest Model:", best_model, "\n")
cat("Best RMSE:", round(results$RMSE[1], 2), "\n")
cat("Best R²:  ", round(results$R2[1], 4), "\n")

cat("\n=== KEY FINDINGS ===\n")
cat("1. Ensemble ML models (XGBoost, Random Forest) outperform parametric models.\n")
cat("2. GAM captures nonlinear effects and provides interpretable smooth terms.\n")
cat("3. Median income is the strongest predictor, but its effect is nonlinear.\n")
cat("4. Location (lat/lon) interacts significantly with income — coastal high-income\n")
cat("   areas show disproportionately higher house values.\n")
cat("5. GNPR methods (GAM, Splines) substantially improve over linear regression\n")
cat("   while remaining more interpretable than black-box ML models.\n")
```




```{r session-info}
sessionInfo()
```
